---
title: "School Of Medicine: R Workshop"
subtitle: "Computer Setup"
author: "University of Adelaide, Bioinformatics Hub"
output: html_document
---

To return to the previous page click [here](https://uofabioinformaticshub.github.io/schoolOfMedicine2016/) or use the back button on your browser.

## R and RStudio

*R* is available from https://www.cran.r-project.org/ 

RStudio is available from https://www.rstudio.com/

Download *R* first, then `RStudio`. 
*R* is the main software we are using, whilst `RStudio` is a user-friendly interface to this software environment.

Please ensure you have a working installation of `RStudio`, simply by opening the software.
We will not need to access *R* directly.

## ADAPT

*R* and `RStudio` are also part of the University's SOE which can be accessed using ADAPT.
If you would prefer to use your desktop on the University's ADAPT system as an alternative to an installation on your own laptop, this should be fine.
However, the code has not been specifically tested in this environment.

__If you plan to use this method, please ensure you can connect and open the software before the workshop.__

## R Setup

Once you have *R* and `RStudio` installed, open `RStudio` and in the `Console` enter the follow commands.
You can cut & paste these as a single command.

```{r, eval=FALSE}
packList <- c("readr", "dplyr", "ggplot2", "readxl", "reshape2")
install.packages(packList)
```

Don't worry too much about all the information that should appear during this process.

### Note for Windows Users

In the above we have installed the package `dplyr`, which can sometimes be problematic on University-owned laptops.
To test that your installation was successful, please cut & copy the following command

```{r, eval = FALSE}
all(packList %in% rownames(installed.packages()))
```

If this returns `TRUE` you are good to go.

If the above returned `FALSE` take the following actions:
    
1. Create a new folder `C:\R`  
2. Create a new `Environment Variable` using the steps below called `R_LIBS_USER` and give it the value `C:\R`
3. After these steps, close then re-open RStudio and run the code again
4. If these steps don't resolve the problem, please connect to your desktop on the ADAPT system and use the version of `RStudio` installed there.

#### Creating a new Environment Variable: Windows 10 and Windows 8

1. In Search, search for and then select: `System (Control Panel)`
2. Click the `Advanced` system settings link.
3. Click `Environment Variables`. 
4. In the User Variables Section, click `New...`
5. Name the New Variable `R_LIBS_USER` and specify the value `C:\R`

#### Creating a new Environment Variable:  Windows 7

1. From the desktop, right click the Computer icon.
2. Choose Properties from the context menu.
3. Click the `Advanced` system settings link.
4. Click `Environment Variables`. 
5. In the User Variables Section, click `New...`
6. Name the New Variable `R_LIBS_USER` and specify the value `C:\R`

[Back](https://uofabioinformaticshub.github.io/schoolOfMedicine2016/)