---
title: "Inspecting Data"
author: "Steve Pederson"
date: "5 December 2015"
output: html_document
---

## Inspecting Data

Once we have our data loaded into R, the first thing we should do is check that it looks right.

The `toothData` file contains data about the effects of Vitamin C on tooth growth in Guinea Pigs, using two delivery methods, and three dose levels.

```{r}
toothData <- read.csv("data/toothData.csv")
head(toothData)
summary(toothData)
nrow(toothData)
```

## Plotting the data

*R* makes visually inspecting the data super-easy, and in particular, the package `ggplot2` has some powerful plotting functions.
If you don't have this package already installed, you can simply install this using the following commands.
This is the simple method for installing packages from the main *R* package repository, known as [CRAN](https://cran.r-project.org/).
A package, is essentially a collection of functions for a some type of analysis.

```{r, eval=FALSE}
install.packages("ggplot2")
```

To then load the functions that ggplot2 contains, we load the entire `library` of functions using.

```{r}
library("ggplot2")
```

```{r}
ggplot(toothData, aes(x = dose, y = len, fill = supp)) +
  geom_boxplot()
```

These need re-ordering...

```{r}
ggplot(toothData, aes(x = len)) +
  geom_histogram()
```

```{r}
ggplot(toothData, aes(x = len, fill = supp)) +
  geom_density(alpha = 0.5)
```


```{r}
ggplot(toothData, aes(x = len, fill = supp)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~dose)
```