---
title: "Introducing Functions"
author: "Steve Pederson"
date: "7 December 2015"
output: html_document
---

## Functions - An Introduction

NB: This section assumes an understanding of what `vector`, `data.frame` and `list` objects are in *R*.
If you're not sure, call a tutor over to discuss this.

Functions are also *R* object, but instead of containing data, they contain instructions for performing some operation.
We use them to "do things", then we usually obtain some output as a plot, or another R object.

- Some well known ones are `max()`, `min()`, `mean()`, `sd()`

```{r, eval=FALSE}
y <- c(1, 16, 2, 9)
mean(y)
min(y)
sd(y)
sqrt(y)
```

- Functions usually take some input, but some special functions don't.

```{r, eval=FALSE}
browseVignettes()
help.start()
Sys.Date()
```

- To execute a function, we need to place round brackets `()` after the name  
(Otherwise we see the code)

**An R Package is essentially just a collection of functions!**

## Passing R objects or arguments to functions

As we've seen, the majority functions require something as the input.
Sometimes, it's a single *R* object, other times it can be multiple objects.
Let's look at the help page for a function.
```{r, eval = FALSE}
?sd
```

Notice that this function requires two arguments, the first is called `x` and the second is called `na.rm`.
These are the names that are essentially just placeholders for the objects we pass to them.

#### The argument `x`

No matter what the object is called that we give to the first position `x`, the function will temporarily create an internal copy of the object, call it `x`, then perform all operations on this object called `x`, leaving your original *R* object unchanged.

When we called the function using `sd(y)` above, the vector `y` was temporarily copied, renamed `x` and then the function operated on this copy of the object.
(NB: This is a slightly crude description for the sake of simplicity, so purists please be understanding)

#### The argument `na.rm = FALSE`

Notice that when we called `sd(y)`, we completely ignored this second argument.
In the function definition, it is given a default value (`na.rm = FALSE`), so we only need to specify this if we want an alternative result.
This argument controls whether we include, or ignore any `NA` values.

#### Specifying arguments by name
In *R*, we can specify arguments stricty by order without naming them, or we can name tham and specify them out of order.
**Which of the following lines of code would be valid?**
```{r, eval=FALSE}
sd(x = y)
sd(y, TRUE)
sd(TRUE, y)
sd(na.rm = TRUE, x = y)
```

We can also quickly find what arguments, and their order, by calling `formals()` on a function.
This will return them as a `list`, along with their default values.
```{r}
formals(sd)
```

## The internal code for a function

As well as arguments, a function will always do something, so some code will be included in every function.
We can see this code, by calling the function `body()`
```{r}
body(sd)
```

This function, only contains one line of code.
Unfortunately, this line may be difficult to understand, so let's write our own in the next section.
