---
title: "Writing Functions"
author: "Steve Pederson & Pengyi Yang"
date: "7 December 2015"
output: html_document
---

## Why would we write functions?

- Repetitive processes

- Nothing may exist for our purpose

- We can customise existing function
    + We may need to tweak the output of a function
    + We may only need part of a larger function
    + We can customise for increased speed

## Our first function

```{r}
myMean <- function(x){
  n <- length(x)
  total <- sum(x)
  return(total / n)
}
```

The functions needs:

1. A `name` so we can 'call' it, and make an educated guess what the function does
2. An `argument` (or input) to 'do something' to.  
3. A `body` which is the code that runs when we call the function


As mentioned previously
 : `x` is just a place holder for "some *R* object"
 : This object then becomes called `x` as far as the function is concerned

Let's run it:

```{r}
y <- c(1, 16, 2, 9)
myMean(y)
```

Check it against the real function

```{r}
mean(y)
```

**NB:** We didn't need to write `myMean(x = y)`.

Just like any other *R* function we can find what the arguments are, and what the code is:
```{r}
formals(myMean)
body(myMean)
```


## Writing functions for NOT you

When we write code, we write for someone who is not you, which is probably you in 6 months, when you have no idea what you did and you think you're an idiot.

To make life easier for our future selves, we can place comments in the code.

```{r}
myMean <- function(x){
  
  # First, count how many values are in x
  n <- length(x)
  
  # Now add them
  total <- sum(x)
  
  # And produce some nice output
  total / n
  
}
```


Notice that

1. any comments are not shown when we call `body()`
2. We didn't include the `return()` command. Functions in *R* will output the final line of code, regardless of whether we include this command.

```{r}
body(myMean)
```

Comments are a great idea for new to intermediate *R* users!

## What could possibly go wrong?

Now we know how to write function that works:  
**How could this ever go wrong?**

Some ways to break `myMean` are:

1. Give it `NA` values
2. Give it something which is not a `numeric` vector

```{r, echo=FALSE, include=FALSE}
ohNo <- c(NA, 1, 3, 5)
```


```{r,eval = FALSE}
ohNo <- c(NA, 1, 3, 5)
myMean(ohNo)
```

## Debugging!

- We might need to debug the function.

- R has an inbuilt function called `browser()`

- If we (temporarily) place this command inside the function, we move
    + **from the Global R Environment**  
    + **to an R Environment INSIDE the function**


```{r}
myMean <- function(x){
  
  browser()
  
  # First, count how many values are in x
  n <- length(x)
  
  # Now add them
  total <- sum(x)
  
  # And produce some nice output
  total / n
  
}
```

Now let's call the function, and a new window will open

You'll also see some new buttons at the top of the R Console

```{r, eval=FALSE}
myMean(ohNo)
```

Turn to the `Environment` Tab at the top right of R Studio.
This is displaying the internal environment of the function

Send the first line in the `Source Viewer` Window to the Console

- `n` will appear in the `Environment` Tab (and so will `x`)

- The value looks right, so this is OK

Now, send the second line to the Console

- The value for `total` is not what we expected, so we need to fix this line!

**Suggestions?**

Let's check the help page to see if there is a clue here as to how we can fix our function.

```{r, eval=FALSE}
?sum
```

A solution might be to use the `na.rm = TRUE` option for the function `sum()`

```{r}
myMean <- function(x){
  
  browser()
  
  # First, count how many values are in x
  n <- length(x)
  
  # Now add the valid numbers
  total <- sum(x, na.rm = TRUE)
  
  # And produce some nice output
  total / n
  
}
```

However, this still wouldn't work as we've still included the `NA` values when we counted them.
My final function would look something like this:
```{r}
myMean <- function(x){
  
  # Check the vector is numeric
  stopifnot(is.numeric(x))
  
  # Remove any NA values
  isNA <- is.na(x)
  x <- x[!isNA]
  
  # Count the valid values
  n <- sum(x)
  
  # Add the valid values
  total <- sum(x)
  
  # And now provide some output
  total / n
  
}
```

