---
title: "Customising Functions"
author: "Steve Pederson & Pengyi Yang"
date: "7 December 2015"
output: html_document
---

## Customising Functions

### The Tooth Growth Data

The `toothData` file actually contains data about the effects of Vitamin C on tooth growth in Guinea Pigs, using two delivery methods, and three dose levels.
The main variable of interest is the length of the teeth, contained in the column `len`.
The delivery was done using two different types of Vitamin C supplements, and these are detailed in the `supp` column.
The dose levels contained in each supplement are then recorded in the column `dose`, and these represent levels of 0.5, 1 and 2.0mg.

```{r}
toothData <- read.csv("data/toothData.csv", stringsAsFactors=FALSE)
toothData$supp <- as.factor(toothData$supp)
toothData$dose <- factor(toothData$dose, levels=c("Low", "Med", "High"))
```

### Data Inspection

In keeping with Section 03: Inspecting Data, let's quickly inspect and plot the data to see what it looks like.

```{r}
head(toothData)
summary(toothData)
nrow(toothData)
```

```{r, fig.height=4}
library(ggplot2)
ggplot(toothData, aes(x = dose, y = len, fill = supp)) +
  geom_boxplot()
```


## Writing A Function For Our Needs

Our task is to conduct a `t.test()` for each dose level, to see if there is any effect on tooth length.
First, let's try it for the low dosage level.
Remember that for a $T$-test, we have the null and alternate hypotheses:

$H_0$: $\mu_1 = \mu_2$

$H_A$: $\mu_1 \neq \mu_2$

Firstly we could do this by breaking the data into two groups and testing the Null Hypothesis:

```{r}
lowData <- subset(toothData, dose == "Low")
splitData <- split(lowData, lowData$supp)
g1 <- splitData$OJ$len
g2 <- splitData$VC$len
t.test(g1, g2, alternative = "two.sided")
```

An alternate method would be to use the *R* syntax, for `len` *is dependent on* `supp`, where the tilde `~` can be interpreted with this meaning.
We can also subset the data as part of the $T$-test function.

```{r}
t.test(len ~ supp, data = toothData, subset = dose == "Low")
```

Doing this for all 3 levels will look messy when we email this to our collaborators! 
Let's see if we can change the structure of the output to make it look nicer.

First, how does *R* really save the results from the `t.test()` function?

```{r}
results <- t.test(len ~ supp, data = toothData, subset = dose == "Low")
str(results)
```

This is an *R* object type called a `list`

*What are we really interested in?*

1. The difference in means?

2. The *T*-statistic

3. The degrees of freedom

4. The *p*-value

**Where are these values in the function output?**

We'll need to calculate the difference in means from the element `$estimate`, but the rest of the objects are in the list elements named `$statistic` `$parameter` & `$p.value`.

We could write a function to output these as a `data.frame`

```{r}
myToothTest <- function(data){
  
  # Perform the t.test
  res <- t.test(len~supp, data)
  
  # Output the data.frame
  data.frame(Difference  = -diff(res$estimate),
             T = res$statistic,
             DF = res$parameter,
             P = res$p.value, 
             row.names= NULL)
  
}
```

Now we can run the function, subsetting the data first.

```{r}
subData <- subset(toothData, dose == "Low")
myToothTest(subData)
```

**This looks much nicer!**

Let's incorporate the sub-setting into the function, and add the dose level to the output.
Note, that some data checking lines have also been introduced to make sure that "Future You", or a collaborator doesn't try something silly.

```{r}
myToothTest <- function(doseLevel, data){
  
  # Check the structure of the inputs
  stopifnot(doseLevel %in% levels(data$dose))
  stopifnot(is.data.frame(data))
  stopifnot(names(data) %in% c("len", "supp", "dose"))
  
  # Subset the data
  data <- subset(data, dose == doseLevel)
  
  # Perform the t.test
  res <- t.test(len~supp, data)
  
  # Output the data.frame
  data.frame(Dose = doseLevel,
             Difference  = -diff(res$estimate),
             T = res$statistic,
             DF = res$parameter,
             P = res$p.value, 
             row.names= NULL)
  
}
```

Now the function is even easier to run:

```{r}
myToothTest("Low", toothData)
```

## Now the FUN starts

*R* has a function called `lapply()` which applies a function to each element of a vector.
As seen on the help page, the results from `lapply()` are as a list, with each element corresponding to the results from each element of the input vector.

First inspect the help page
```{r, eval=FALSE}
?lapply
```

As a trivial example, just to see this process in action:
```{r}
y <- c(1, 16, 2, 9)
lapply(y, FUN = sqrt)
```

**This places each element in the FIRST argument position of the function**.  
First, `lapply()` placed `1` into the first position of the function `sqrt()`.
Then it passed `16`, then `2` followed by `9`, and each result was output as a individual element of a `list()` of length 4.

There is a alternate function `sapply()` that simplifies the output where possible

```{r}
sapply(y, FUN = sqrt)
```

It's very useful, but slightly unpredictable, so use with caution in long, automated procedures or functions.

**Can we use lapply for our different dose levels?**

YES!!!
This is why we placed `doseLevels` as the first argument of our function

```{r}
doseLevels <- levels(toothData$dose)
lapply(doseLevels, myToothTest, data = toothData)
```

Note that we specified `data = toothData` after the function call.
This is where we place any other arguments to the function being called.

## Getting amazing looking output

The *R* package `dplyr` has a function `bind_rows()` for joining identical `data.frame` objects.

First install and/or load the package
```{r, eval=FALSE}
install.packages(dplyr)
library(dplyr)
```

```{r, echo=FALSE, message=FALSE}
library(dplyr)
```

```{r, warning=FALSE}
doseResults <- lapply(doseLevels, myToothTest, data = toothData)
doseResults <- bind_rows(doseResults)
doseResults
```

And for those using R Markdown/knitr

```{r, warning=FALSE}
library(knitr)
kable(doseResults)
```

## In the real world

This technique can be applied to numerous datasets

- You might want to run the same linear regression model on data from different climates/regions

- You might have thousands of SNPs to run a `fisher.test()` on

- The *R* package `parallel` also has a version of this called `mclapply()` which runs the process in parallel.
Whilst there is some overhead as *R* configures each CPU core as an internal node on a cluster, this can greatly speed up results for long functions, or large datasets.
The syntax is identical to `lapply()`, making this step-up very simple.  
