---
title: "The R Package clueR"
author: "Pengyi Yang & Steve Pederson"
date: "7 December 2015"
output: html_document
---

## Reading data in R and exploratory data analysis

#### Read in a tab-delimited dataset from plain text file and calculate summary statistics
```{r}
Adipo.dat <- read.delim("data/Adipogenesis.txt", head=TRUE)
Adipo.dat[1:5,]
```

subsetting a matrix
```{r}
Adipo.exp <- as.matrix(Adipo.dat[,-1]) # removing the ID column from the raw data to create a numeric matrix containing the expression of the genes at each time point
rownames(Adipo.exp) <- Adipo.dat[,1] # assign the row names of the matrix as the ID in the first column
Adipo.exp[1:5,] # displaying the first five rows of the expression matrix
summary(Adipo.exp)
```

Read data from R data object
```{r}
load("data/DEGs.RData") # This data contains differentially expressed genes
# subset the expression matrix to select only DE gene experssions
Adip.exp.DEGs <- Adipo.exp[DEGs,]
```

write table
```{r}
#write.table(Adip.exp.DEGs, file="Adipogenesis_DE.txt", quote=F, sep="\t")
#save(Adipo.dat, file="Adipo.RData")
```


#### Exploratory data analysis and visualisation
boxplot
```{r}
par(mfrow=c(1, 2))
boxplot(Adipo.exp, ylab="log2 Fold Change", main="Full dataset")
boxplot(Adip.exp.DEGs, ylab="log2 Fold Change", main="DE dataset")
```


```{r}
par(mfrow=c(1, 2))
# Scatter plot the full data and highlight the DE genes in the scatter plot
plot(Adipo.exp[,c("t8h", "t48h")])
points(Adipo.exp[DEGs,c("t8h", "t48h")], pch="*", col="red", main="8h vs 48h")
abline(h=0, v=0, col="gold")
# generating pairwise correlation plots
plot(Adipo.exp[,c("t4h", "t48h")])
points(Adipo.exp[DEGs,c("t4h", "t48h")], pch="*", col="red", main="4h vs 48h")
abline(h=0, v=0, col="gold")
```




## Install package and calling functions from packages
Install an R package called 'corrplot' from CRAN (The Comprehensive R Archive Network)
```{r}
install.packages("corrplot", repos='http://cran.us.r-project.org')
library(corrplot)
#?corrplot
corrplot(corr=cor(Adip.exp.DEGs, use="na.or.complete"), addCoef.col="red")
```


### Practice 2
Cluster evaluation R package (clueR) is a package to determine optimal fuzzy k-mean clustering from time-series data. Please following the steps below to analyse and visualise the time-series data from Adipogenesis:

**Step 1:** Install an R package called 'clueR' from CRAN.
```{r}
install.packages("ClueR", repos='http://cran.us.r-project.org')
```

**Step 2:** Run the following two lines of code in R.
```{r}
library(ClueR)
data(Pathways)
rownames(Adip.exp.DEGs) <- toupper(rownames(Adip.exp.DEGs))
clueObj <- runClue(Tc=Adip.exp.DEGs, annotation=Pathways.KEGG, rep = 5, kRange=20)
```

**Step 3:** clueObj is a list that contains various evaluation information and statistics. clueObj$evlMat is the evaluation matrix that contains the clustering goodness from k=2 to k=20. Calculate means and standard deviations with respect to each column of this matrix.

**Step 4:** Install an R package called 'Hmisc', plot means calculated in step 3, and add standard deviations as error bar using function 'errbar' from 'Hmisc'.
```{r}
Ms <- apply(clueObj$evlMat, 2, mean, na.rm=TRUE)
Ss <- apply(clueObj$evlMat, 2, sd, na.rm=TRUE)
library(Hmisc)
errbar(1:length(Ms), Ms, Ms+Ss, Ms-Ss, cex=1.2, type="b", xaxt="n", xlab="k", ylab="E")
axis(1, at=1:12, labels=paste("k=", 2:13, sep=""))
```

**Step 5:** Run the following code to obtain optimal clustering result determined by clueR
```{r}
# generate the optimal clustering results
 best <- clustOptimal(clueObj, rep=10, mfrow=c(4, 4))
# list enriched clusters
 best$enrichList
```




